\documentclass[12pt]{article}

\usepackage{titlesec}
\usepackage{enumerate}
\usepackage[parfill]{parskip}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{graphicx}
\usepackage{verbatim}
\usepackage{listings}

\usepackage{xcolor}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

\lstdefinestyle{mystyle}{
    language=Python,
    backgroundcolor=\color{backcolour},   
    commentstyle=\color{codegray},
    keywordstyle=\color{codegreen},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\scriptsize,
    breakatwhitespace=false,         
    captionpos=b,                    
    keepspaces=true,                 
    numbers=left,                    
    numbersep=5pt,                  
    showspaces=false,                
    showstringspaces=false,
    showtabs=false,                  
    tabsize=2
}

\lstset{style=mystyle}


\titleformat{\section}
    {\normalfont\fontsize{15}{17}\bfseries}{\thesection}{2.27em}{}
\titleformat{\subsection}
    {\normalfont\fontsize{17}{18}\bfseries}{\thesubsection}{1em}{}
\titleformat{\subsubsection}
    {\normalfont\fontsize{15}{15}\bfseries}{\arabic{subsubsection}}{1em}{}

\newcommand{\starreditem}{\item[\refstepcounter{enumi}\number\value{enumi}*.]}
\newcommand{\textmod}[1]{\;(\text{mod }#1)}
\newcommand{\mathmod}[1]{\quad(\text{mod }#1)}

\begin{document}
\setcounter{section}{14}
\section{Number Theory}
\setcounter{subsection}{5}
\subsection{Computing roots Modulo \textit{p}}
\emph{CATAM coursework for Part II of the Mathematical Tripos. Sections have
been numbered as they appear in the manual.}

\setcounter{subsubsection}{1}
\subsubsection{Computing Legendre Symbols}
\textbf{Question 1} \quad 
We implement the repeated squaring method for modular exponentiation using a
recursive algorithm: 
\[a^n \equiv \begin{cases}
    (a^{\frac{n}{2}})^2, &n\text{ even}, n>0\\
    a\cdot (a^{\frac{n-1}{2}})^2, &n\text{ odd}\\
    1, &n = 0
\end{cases} \quad (\text{mod } p)\]
This allows for efficient application of Euler's criterion to compute Legendre
symbols. For \(p=30275233\), we compute \((a/p)\) for \(a\) taking:
\begin{enumerate}[(i)]
    \item \(100\) random values between \(1\) and \(p\). Out of these, 52
        numbers were found to be quadratic residues mod \(p\).
    \item all values between \(1\) and \(100\). Out of these, \(58\) numbers
        were found to be quadratic residues mod \(p\).
\end{enumerate}
Appendix A contains a record of all output produced. Note that in both the
cases, approximately half of the numbers tested turned out to be quadratic
residues. This suggests that one might model quadratic residues in a finite
field using a uniform distribution.

 \hfill

\textbf{Question 2}\quad Suppose \(n\) is any odd number and \(m\in
\mathbb{Z}\).  If \(n=1\), we have \((m/n)=(m/1)=1\). Otherwise, we may assume
\(0\leq m < n\) by reducing mod \(n\) if necessary. Note that \((0/n)=0\) when
\(n>1\). If \(m\) is a non-zero even number, we may get rid of
all factors of \(2\) using the following property of the Jacobi symbol:

\textbf{Lemma} For any positive odd \(n\) we have
\[\left(\frac{2}{n}\right) = (-1)^{(n^2-1)/8}= \begin{cases} +1, &n \equiv \pm1
    \mathmod{8} \\ 
-1, &n\equiv \pm 3 \mathmod{8}\end{cases}\]

A proof of this can be found in [1] (\textit{pp.} 47, Proposition
II.2.6). Hence we may assume \(m,n\) are both odd and \(1\leq m<n\). We state the following strengthening of quadratic reciprocity: 

\textbf{Lemma} For any two positive odd integers \(m\) and \(n\) we have
\[\left(\frac{m}{n}\right) = (-1)^{(m-1)(n-1)/4}\left(\frac{n}{m}\right).\]

A proof of this can be found in [1] (\textit{pp.} 47, Proposition
II.2.7), but it follows immediately from the law of quadratic reciprocity and
the observation that \((m/n)=0\) whenever \(m\) and \(n\) are not coprime.
Moreover, we can avoid computing the large product \((m-1)(n-1)/4\) by observing
that 
\[(-1)^{(m-1)(n-1)/4} = \begin{cases}
    -1, & m,n\equiv 3 \quad(\text{mod }4) \\
    +1, &\text{otherwise}
\end{cases}\] 

To compute \((m/n)\), it now suffices to compute \((n/m)\). Note that the
`denominator' strictly decreases after each iteration, hence the recursion must
halt in finite time.

We compute the complexity of our algorithm which computes \((m/n)\), without
loss of generality \(m\geq n\). If the numerator at each step is odd, the
algorithm operates like Euclid's algorithm for the greatest common divisor. In
particular, there are at most \(O(\log n)\) reductions of form \((m/n)\mapsto (n
\mod m / m)\). However, with each reduction we also remove all powers of \(2\)
from the numerator. Since the numerator is bounded by \(m\), each reduction
takes \(O(\log m)\) basic operations hence the entire algorithm has complexity
\(O(\log m \log n)\).

Reducing mod \(p\), we may assume \(0\leq m,n < p\) so that the algorithm has
complexity \(O((\log p)^2)\). In contrast, the repeated squaring algorithm of
Question 1 computes the Legendre symbol in \(O(\log p)\) basic operations. While
it may seem that repeated squaring is faster, we must also acknowledge the
fact that the `basic' operations involved are multiplications of numbers
containing \(\log p\) bits. For large \(p\), this significantly affects the
performance of the algorithm from Question 1 whereas the second algorithm
remains relatively unaffected.

\subsubsection{Computing square roots mod \textit{p}}
\textbf{Question 3}\quad 
Suppose \(p\equiv 3\;(\text{mod }4)\) and \(a\) is a quadratic residue mod \(p\). In
particular, Euler's criterion implies \(a^{(p-1)/2}\equiv 1\textmod{p}\). Then \(x\equiv
a^{(p+1)/4}\) is a solution to \(x^2\equiv a\textmod{p}\) since
\[x^2 \equiv a^{(p+1)/2} \equiv a\cdot a^{(p-1)/2} \equiv a \mathmod{p}.\]

Suppose \(p\equiv 5\textmod{8}\). We have \((2/p) \equiv
2^{(p-1)/2}\equiv -1 \textmod{p}\). If \(a\) is a quadratic residue, we have
\(a^{(p-1)/2}\equiv 1 \Rightarrow a^{(p-1)/4}\equiv \pm 1 \textmod{p}\). In
particular, we can write \( 2^{k(p-1)/2} \cdot a^{(p-1)/4}\equiv 1 \textmod{p}\) for some \(k\in
\{0,1\}\). But then we have \(a \equiv 2^{k(p-1)/2} \cdot a^{(p-1)/4}\cdot a \equiv
2^{k(p-1)/2}\cdot a^{(p+3)/4} \textmod{p}\). Observe that all the exponents
involved are even, so we can read off a solution to \(x^2\equiv a\) as \(x\equiv
2^{k(p-1)/4}\cdot a^{(p+3)/8}\textmod{p}\).

\hfill 

Now suppose \(p\) is a prime of the form \(2^n+1\) (we only consider \(n>2\)
since \(p=3,5\) have been covered above). Then \(p\equiv (-1)^n+1\textmod{3}\),
hence \(n\) must be even. It follows that \(p\equiv 1\textmod{4}\) and \(p\equiv
2\textmod{3}\), so that \((3/p) = (p/3) = (2/3) = -1\). Let \(g\) be any
primitive root mod \(p\). The subgroup \(\langle g^2\rangle\) has order
\(2^{n-1}=\frac{p-1}{2}\) hence contains all the quadratic residues. Moreover,
it is the unique multiplicative subgroup of order \(2^{n-1}\). Since \(3\notin
\langle g^2\rangle\), the multiplicative order of \(3\) must be \(2^n\) i.e.\
\(3\) is also a primitive root mod \(p\). 

\textbf{Question 4}\quad
Suppose \(p=65537=2^{16}+1\), and we wish to solve the congruence \(x^2\equiv
18612 \textmod{p}\). We compute \((18612/65537)=1\) using the program written
for Question 2, so such an \(x\) exists. For the purposes of this question, use
`\(=\)' to denote congruence mod \(p\). For modular exponentiation, we use a
program based on the repeated squaring method.

Since \(3\) is a primitive root mod \(p\), we may write \(x =
3^{r_0+2r_1+2^2r_2+...}\) where each \(r_j\in \{0,1\}\). The congruence
\(x^2=18612\) can be written as \(\prod_{j\geq 0} 3^{r_j2^{j+1}} = 18612
\textmod{p}\). 

Raising both the sides to \(2^{14}\), all the terms for \(j\geq 1\) vanish and
we are left with \(3^{r_02^{15}}= 18612^{2^{14}} = 1\). Since \(3\) is a
primitive root, \(3^{2^{15}} = -1\) by Euler's criterion and we have \(r_0=0\).
In fact, \(18612^{2^n}=1\) for all \(14\geq n\geq 11\) hence we have \(r_0=...=r_3 =
0\). 

Raising both the sides of \(\prod_{j\geq 4}3^{r_j2^{j+1}}=18612\) to the power
\(2^{10}\), we obtain \(3^{r_42^{15}}=-1\), hence \(r_4 = 1\). We may multiply
both the sides of the congruence by \(3^{-r_42^5} = 3^{2^{16}-2^5} = 29606\) to
obtain \(\prod_{j\geq 5}3^{r_j2^{j+1}}=57313.\) 

Again, \(57313^{2^n}=1\) for \(9\geq n \geq 6\), so \(r_5=...=r_8=0\), and we
have \(\sum_{j\geq 9}3^{r_j2^{j+1}} = 57313\). Raising both the sides to the
power \(2^5\), we obtain \(3^{r_92^{15}}=-1\), hence \(r_9 = 1\). Multiply both
the sides by \(3^{-r_92^{10}} = 3^{2^{16}-2^{10}} = 64509\) to obtain
\(\prod_{j\geq 10}3^{r_j2^j} = 65536 = -1\). Comparing with \(3^{2^{15}} = -1\),
we deduce \(r_{10} = ... = r_{13}=0\) and \(r_{14}=1\).

We can now read off the solution to \(x^2 = 18612\) as \(x= 3^r\) where
\(r=2^4+2^9+2^{14}\). A square root of \(18612\) (mod \(p\)) hence is
\(45462\). The other square root is \(-45462= 20075\) and these are the only
solutions to \(x^2 =18612\) since \((\mathbb{Z}/p\mathbb{Z})^\ast\) is a field.

\hfill 

Suppose \(p\) is any odd prime and \(a\) is a quadratic residue mod \(p\). In
this section, use `\(=\)' to denote congruence mod \(p\). We may find \(\alpha >
0\) and \(s\) odd such that \(p-1 = 2^\alpha s\). Since \(s\) is odd, we
define \(z = a^{(s+1)/2}\) and observe that if \(y^2 = a^s\), then \(zy^{-1}\)
is a square root of \(a\) mod \(p\) (where \(y^{-1}\) is the multiplicative inverse of
\(y\) in \((\mathbb{Z}/p\mathbb{Z})^\times\).)

Now \((a^s)^{2^{\alpha - 1}} = a^{(p-1)/2} = (a/p)= 1\), so \(y\) is an  element of the
cyclic multiplicative group \(G=\{g\in
(\mathbb{Z}/p\mathbb{Z})^\times\;|\;g^{2^{\alpha}} = 1\}\). Suppose \(n\) is any
non-residue mod \(p\), and let \(b=n^s\). Then we have \(b^{2^\alpha}=n^{p-1} =
1\) hence \(b\in G\), and moreover \(b\) generates the group since
\(b^{2^{\alpha - 1}}=n^{(p-1)/2} = (n/p)=-1\). We can then write \(y = b^r\),
and solve for \(r\) algorithmically by considering its binary expansion. A
square root of \(a\), then, is \(zb^{2^{\alpha}-r}\).

\textbf{Question 5}\quad We implement an algorithm that uses the above method to
compute square roots mod \(p\) when \(p\equiv 1 \textmod{8}\), using the more
direct computations from Question 3 to handle other cases. The second solution
to the congruence \(x^2\equiv a \textmod{p}\) can be computed as \(-1\) times the first
solution. 

Here are some test cases, in the form \texttt{(a, root a)}. The values of \(a\)
(other than the first) have been chosen randomly subject to being quadratic
residues. The primes have been chosen\footnote{The first prime is the largest
known Fermat prime, while the latter four are relatively large palindromic
primes [OEIS: A055578].} to cover all possible congruence classes mod \(8\). The
generated roots can be confirmed to be accurate by squaring them. In particular,
the first test case agrees with what we found in Question 4.

\verbatiminput{../output/Test-roots.txt}

Additionally, we compute the roots of all quadratic residues in
\(\{1,2,...,20\}\) mod \(30275233\) and present the results in Appendix B.

The complexity of this algorithm can two sources, the first of which is finding
\(b\). We run an exhaustive search, computing \((n/p)\) for \(n=1,2,...\) until
a non-residue is hit. In the worst case this has complexity \(O(p(\log p)^2)\),
but assuming that quadratic residues are uniformly distributed in
\((\mathbb{Z}/p\mathbb{Z})^\times\) i.e.\ if \(\mathbb{P}((a/p)=1) =
\frac{1}{2}\), we expect to succeed in \(\sum_{n=1}^\infty n (\frac{1}{2})^n =
2\) steps. Hence this part of the algorithm effectively has complexity
\(O(1)\) (since \(n\approx 1\)). 

Hence all of the complexity comes from computing \(r\), which we can bound by
\(p\). Then we need to compute the \(O(\log p)\) digits of \(r\), each of
which takes \(O(\log p)\) steps to compute (since the computation involves
exponentiation). Hence we estimate the entire algorithm to have complexity
\(O((\log p)^2)\).

\subsubsection{Computing roots of polynomials mod \textit{p}}
Let \(p\) be a prime.

\textbf{Question 6}\quad Given two polynomials \(f\) and \(g\) in
\((\mathbb{Z}/p\mathbb{Z})[x]\), note that \(\mathbb{Z}/p\mathbb{Z}\) is a field
hence the leading coefficient of \(g\) is a unit-- moreover, we can use Fermat's
little theorem to explicitly compute its inverse using the repeated squaring
method. With this, we implement a recursive algorithm to eliminate terms in
\(f\) of degree higher than \(\deg(g)\) and find polynomials \(q,r\) such that
\(f = qg+r\), \(\deg(r)<\deg(g)\). Write \(r=\text{rem}(f,g)\). 

This allows for the application of Euclid's algorithm to find the greatest
common divisor of \(f\) and \(g\), by 
\[\text{gcd}(f,g) = \begin{cases} f, &g=0 \\
\text{gcd}(g,\text{rem}(f,g)),&\text{otherwise}\end{cases}.\]
The answer is generated up to a unit, so we divide by the leading coefficient to
normalise. To test the algorithm, we compute the greatest common divisors for
the following pairs of polynomials: 
\begin{align*}
    &\text{gcd}(x^3+6x^2+5x+5, x^3+13x^2+6x+3) = x^2 + 78x + 62 &(p=109). \\
    &\text{gcd}(x^3+2x^2+9x+4, x^3+3x^2+7x+9) = 1 &(p=131). \\
    &\text{gcd}(x^3+3x^2+9x+12, x^3+6x^2+12x+4) = x+83 &(p=157). \\
\end{align*}

If \(f,g,h\) are polynomials, write \(f\equiv g\textmod{h}\) to mean `there
exists a polynomial \(q\) such that \(f = qh+g\).' It is straightforward to
check that this is an equivalence relation, and \(f_1\equiv g_1\textmod{h}\),
\(f_2\equiv g_2\textmod{h}\) implies \(f_1+f_2 \equiv g_1+g_2\textmod{h}\) and
\(f_1f_2\equiv g_1g_2\textmod{h}\).  Moreover, \(\text{rem}(f,h)\) is the unique
\(g\) such that \(f\equiv g\textmod{h}\), \(\deg(g)<\deg(h)\). It follows that
\(f\equiv f'\textmod{h}\) implies \(\text{rem}(f,h)=\text{rem}(f',h)\).

\textbf{Question 7}\quad Let \(f \in (\mathbb{Z}/p\mathbb{Z})[x]\) be any polynomial. The polynomial
\(\Phi(x)=x^p-x \) factorises as \(\Phi(x) = \prod_{i=0}^{p-1}(x-i)\) hence
\(g=\text{gcd}(f,\Phi)\) is a product of distinct linear factors. Moreover,
\((x-i)\;|\;f\) if and only if \((x-i)\;|\;g\), hence to compute the roots of
\(f\) it suffices to compute the roots of \(g\).

Now the computation of \(g\) using Euclidean algorithm takes at most \(\deg(f)\)
steps, and in each step except for the first one, the degree of polynomials
involved is less than \(\deg(f)\). Naively computing \(\text{rem}(\Phi, f)\)
would take \(O(p)\) steps; however we observe that
\[\text{rem}(x^p-x,f)\equiv \text{rem}(x^p,f)-\text{rem}(x,f) \mathmod{h}\]
The second term is straightforward to compute. For the first term, we can use a
repeated squaring algorithm similar to that for modular exponentiation of
numbers. This speeds up the first step of Euclidean algorithm to take \(O(\log
p)\) steps, and moreover the polynomials involved in each computation have
degree at most \(2\deg(f)\).

Having reduced \(f\) to the case where it is a product of distinct linear
factors, we fix a \(v \in \mathbb{Z}/p\mathbb{Z}\) and compute
\(\text{gcd}((x+v)^{(p-1)/2}-1, f)\). Everything that has been said about
avoiding large powers in the computation applies here. The greatest common
divisor is \(f\) if \(\alpha+v\) is a quadratic residue for every root
\(\alpha\) of \(f\), it is \(1\) if \(\alpha+v\) is a non-residue for every root
\(\alpha\) of \(f\). Otherwise, we have arrived at a non-trivial factor
of \(f\), and can repeat the process with these till we have found all
linear factors (and hence the roots).

How many tries will this algorithm require before successfully factorizing a
quadratic? For simplicity assume that quadratic residues are uniformly
distributed in
\((\mathbb{Z}/p\mathbb{Z})^\times\), i.e.\ \(\mathbb{P}((n/p)=1)=\frac{1}{2}\)
for \(n\) sampled randomly. Then in this simplified model, the
probability of obtaining a successful factorisation \((x-\alpha)(x-\beta)\)
with a randomly chosen \(v\) is \(\frac{1}{2}\) (i.e.\ probability that both \(v+\alpha\) and \(v+\beta\) are residues or non-residues). Hence the expected number of tries before success is \(\sum_{n=1}^\infty n (\frac{1}{2})^n=2\).

We employ this method to find square roots by using \(f=x^2-a\). The
roots of all quadratic residues between \(21\) and \(99\) are computed
and recorded in Appendix B. The averate number of tries until success
is \(1.83\) which is very close to the expected value predicted by our model.

Given the discussion above, for calculation of complexity we assume the
algorithm succeeds on the first trial. Since \(f\) is quadratic, the complexity
of polynomial division arises from having to invert leading coefficients which
takes \(O(\log p)\) steps. The faster algorithm for computing greatest common
divisors takes \(O(\log p)\) steps to compute
\(\text{gcd}((x+v)^\frac{p-1}{2}-1, f)\) hence the entire algorithm has
complexity \(O(\log p)\). This is significantly faster than
the algorithm of Question 5 for large \(p\).

\textbf{Question 8} \quad We use the algorithm discussed above to compute the
roots of the following biquadratic polynomials modulo \(35564117\):

\begin{tabular}{ll}
    \(x^4 + 5x^3 + 12x^2 + 6\) &\(\{7174009, 9335487, 21485344, 33133389\}\)\\
    \(x^4 + x^3 + 3x^2 + 7x + 4\) &\(\{22043805, 13520313, 35564116\}\)\\
    \(x^4 + 4x^3 + 15x^2 + 3x + 8\) &\(\{16498240, 3842901\}\)
\end{tabular}

\subsubsection*{Appendix A: Computed Legendre symbols}
Using Euler's criterion, Legendre symbols were computed for \(100\) random
values between \(1\) and \(p=30275233\). The results are recorded here in the
format \(\texttt{a: (a/p)}\).

\verbatiminput{../output/Legendre-random.txt} 

We also perform a similar calculation for all \(a\) between \(1\) and \(100\).

\verbatiminput{../output/Legendre-100.txt} 

\subsubsection*{Appendix B: Roots of residues mod 30275233}
For every integer \(1\leq a\leq 20\), we check if \(a\) is a quadratic residue
mod \(p=30275233\) and if so, compute a solution to \(x^2\equiv a\textmod{p}\)
using the method of Question 5.
The computed roots are presented in format \texttt{(a, root a)}. Note that the second root is simply the additive inverse of the given one.

\verbatiminput{../output/Twenty-roots.txt} 

We perform a similar exercise for integers \(21 \leq a < 100\), this time
using the method of Question 7 instead i.e.\ computing the roots of \(x^2-a\). 
The computed square roots are presented in format \texttt{(a, root a)}. We also
record the average number of values of \(v\) used before arriving at a
successful factorisation.

\verbatiminput{../output/Twenty-roots-2.txt} 

\pagebreak

\subsubsection*{Appendix C: Programs}

\underline{\texttt{jacobi.py}}
\lstinputlisting{../src/jacobi.py}

\underline{\texttt{sqroots.py}}
\lstinputlisting{../src/sqroots.py}

\underline{\texttt{polyroots.py}}
\lstinputlisting{../src/polyroots.py}

\underline{\texttt{main.py}}
\lstinputlisting{../src/main.py}


\subsubsection*{References}
\begin{enumerate}[{[}1{]}]
    \item Koblitz, N. \textit{A course in Number Theory and Cryptography},
        Graduate Texts in Mathematics 114, Springer, 1987.
\end{enumerate}
\end{document}
